Addition and removal energies of circular quantum dots

### 2017-06-13 Manuscript/paper.tex
Replace ~ with \<space>

In this work we will mainly work with the formalism of second quantization, see for example Ref.~\cite{shavitt2009many}.  To properly describe fermions, we introduce the annihilation and creation operators $\hat a_p$ and $\hat a_p^\dagger$. These operators must satisfy the anticommutation relations that encode the Pauli principle, namely
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}.
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.~\eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.  In this paper, we use the letters $p, q, r, s, t, \ldots$ to label arbitrary single-particle \textit{states} in the basis.  For the harmonic oscillator basis, labels such as $p$ can be considered abbreviations for triplets of the form $(n, m_\ell, m_{\mathrm{s}})$.  A summation such as $\sum_p$ is then understood as a summation over every single-particle basis state $p$.  This is not to be confused with summations such as $\sum_j$ from \eqref{eq:onetwobodyhamiltonian}, which contains a label $j$ denoting an arbitrary \textit{particle}.  We thus see another benefit with second quantization: the Hamiltonian representation makes no attempt to distinguish particles at all, as expected for particles with quantum statistics.

@@

Aside from trivial cases, the HF equation is generally solved using an iterative algorithm.  We begin with an initial guess $\bm{C}^{(i)}$, which is fed into Eq.~\eqref{eq:fock} to produce the Fock matrix.  This is then used in Eq.~\eqref{eq:hartreefock}, which leads to a standard eigenvalue problem from which $\bm{C}^{(i + 1)}$ arise as the matrix of eigenvectors and $\bm{\varepsilon}^{(i + 1)}$ as the vector of eigenvalues.  This process can be repeated indefinitely until $\bm{C}$ approaches a fixed point.  While in theory it is possible for the solution to never reach a fixed point, or that it may require a unfeasibly large number of iterations, in practice this naive approach can adequately provide solutions for many cases.  In other cases where it is insufficient, methods such as DIIS, Broyden's method, or even \textit{ad hoc} linear mixing can improve and accelerate convergence greatly.  Therefore, the possibility of slow or non-convergence is generally not a concern in practice.

@@

To make use of the HF solution as the reference state for future calculations, we transform the operators via Eqs.~\eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  Therefore, the post-HF methods described in the next few sections generally use the transformed operators $\hat{H}^{[1] \prime}$ and $\hat{H}^{[2] \prime}$ as inputs.  However, we will not annotate them with the prime symbols as the methods are generic and can be applied to any reference state, whether optimized by HF or not.

@@

The central theme of similarity renormalization group (SRG) methods is the application of a continuous sequence of unitary transformations on the Hamiltonian to evolve it into a band- or block-diagonal form, see for examples Refs.~\cite{kehreinflow2006,heikophysrep2016,lnp936} for derivations and calculational details.

@@

The sequence of transformations is parameterized by a continuous variable $s$ known as the \textit{flow parameter}.  Without loss of generality, we can define $s = 0$ to be the beginning of this sequence, thus $\hat{H}(0)$ is simply the original Hamiltonian.  At any value of $s$, the evolving Hamiltonian $\hat{H}(s)$ is related to the original Hamiltonian by
\begin{align*}
  \hat{H}(s) \equiv \hat{U}(s) \hat{H}(0) \hat{U}^\dagger(s),
\end{align*}
where $U(s)$ is a unitary operator that describes the product of all such transformations since $s = 0$.  Taking the derivative with respect to $s$, we obtain:
\begin{align*}
  \frac{\D}{\D s} \hat{H}(s) = \frac{\D \hat{U}(s)}{\Ds} \hat{H}(0) \hat{U}^\dagger(s) + \hat{U}(s) \hat{H}(0) \frac{\D \hat{U}^\dagger (s)}{\D s}.
\end{align*}
If we define an operator $\hat{\eta}(s)$ as
\begin{align} \label{eq:etadefinition}
  \hat{\eta}(s) \equiv \frac{\D \hat{U}(s)}{\D s} \hat{U}^\dagger(s),
\end{align}
we find that it is antihermitian as a result of the unitarity of $\hat{U}(s)$:
\begin{align*}
  \hat{\eta}(s) + \hat{\eta}^\dagger(s)
  = \frac{\D}{\D s} \left(\hat{U}(s) \hat{U}^\dagger(s)\right)
  = 0,
\end{align*}
From this property we can derive a differential equation known as the \textit{SRG flow equation}:
\begin{align} \label{eq:imsrgode}
  \frac{\D \hat{H}(s)}{\D s} = [\hat{\eta}(s), \hat{H}(s)].
\end{align}
This equation allows $\hat{H}(s)$ to be evaluated without explicitly constructing the full transformation $\hat U(s)$.  The focus is instead shifted to the operator $\hat \eta(s)$, the \textit{generator} of the transformation.  When $\hat \eta(s)$ is \emph{multiplicatively} integrated (\textit{product integral}), the full unitary transformation $\hat U(s)$ is recovered:
\begin{align} \label{eq:etaintegral}
  \hat U(s')
  = \lim_{\Delta s \to 0} \prod_{i = 1}^n \mathrm{e}^{\hat{\eta}(s_i) \Delta s},
\end{align}
where $s_i \equiv i \Delta s$, $n \equiv \lfloor s' / \Delta s \rfloor$, and $\lfloor x \rfloor$ denotes the \textit{floor} of $x$.  This is the formal solution to the linear differential equation Eq.~(\ref{eq:etadefinition}).  The product integral in Eq.~(\ref{eq:etaintegral}) may also be reinterpreted as \textit{$s$-ordering} \cite{reimann2013quantum} in analogy to time-ordering from quantum field theory.

@@

The SRG flow equation Eq.~\eqref{eq:imsrgode} can be solved in the second quantization formalism described in Section \ref{subsec:modelHamiltonian}, where field operators are defined with respect to the physical vacuum state.  However, since the basis of a many-body problem grows factorially with the number of particles and the size of the model space, the applicability of the naive (free-space) SRG method is restricted to comparatively small systems.  A more practical approach is to perform the evolution \textit{in medium}, i.e.\ using a many-body Slater determinant as a reference \cite{kehrein2006flow}.  This gives rise to the IM-SRG method \cite{bogner2012,heikophysrep2016,lnp936}.

The second-quantized operators in Eq.~\eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1, i_2, \ldots, i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators.  The normal ordering of the quasiparticle field operators is straightforward to compute.  What is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $|\Phi\rangle$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &=
  \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p, q}^{}.
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_{i} n_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} n_i n_j \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_{i} n_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix $F_{p q}$.  The operator $\hat H$ in Eq.~\eqref{eq:normordhamiltonian} is completely equivalent to $\hat H$ in Eq.~\eqref{eq:vacuumhamiltonian}.  The only difference is that the meaning of a ``$k$-body operator'' has been redefined with respect to the reference state $\Phi$ rather than vacuum state $\ket{}$, causing matrix elements to be reshuffled among the $k$-body operators.  Where this makes a critical difference is when the operator expressions are \emph{truncated}, i.e.\ some $k$-body operators may be discarded from the computation for efficiency reasons.  Normal-ordering can drastically alter the importance of the $k$-body operators with respect to each other.

@@

By expanding the commutator in Eq.~\eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H) \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \mathcal{A}_{A, B} D^{[k]}(A, B) \text{ for any $k$, $A$, $B$}, \label{eq:commutmxe} \\
  \mathcal{A}_{x, y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr) \text{ for any $x$, $y$, $f$},
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} n_i \bar{n}_a A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} n_i n_j \bar{n}_a \bar{n}_b A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p, q}(A, B)
  &\equiv
    \sum_{r} A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} n_i n_j \bar{n}_a A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} n_i \bar{n}_a \bar{n}_b A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} n_i \bar{n}_a \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right)
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    \tilde{\mathcal{A}}_{p, q} \tilde{\mathcal{A}}_{r s}
    \sum_{i a} n_i \bar{n}_a A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    \notag \\
  &\hphantom{=}
    + \frac{1}{2} \sum_{i j} n_i n_j A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} \bar{n}_a \bar{n}_b A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + \sum_{t} \left(
    \tilde{\mathcal{A}}_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \tilde{\mathcal{A}}_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}.
    \right)
    \label{eq:flow2}
\end{align}

\begin{figure}
\includegraphics{fig-diagrams-imsrg.pdf}
\caption{Diagrammatic representation of $D^{[k]}(\circ, \bullet)$ of Eq.~\eqref{eq:flow0}, Eq.~\eqref{eq:flow1}, and Eq.~\eqref{eq:flow2}, where open circles represent $A$ and filled circles represent $B$.  The two-body vertices are implicitly antisymmetric (i.e.\ \textit{Hugenholtz diagrams} \cite{HUGENHOLTZ1957481}, as described in \cite{shavitt2009many} \S 4.4.3).  Internal directed lines (bound/dummy indices) that point upward denote summations over unoccupied states, whereas lines that point downward denote summations over occupied states.  External undirected lines denote unconstrained free indices.  In IM-SRG(2), the three-body diagrams are not included.}
\label{fig:diagrams-imsrg}
\end{figure}

The equations can be derived directly from the anticommutation relations of the field operators, but it is more convenient to use Wick's theorem \cite{PhysRev.80.268} or, even more efficiently, diagrammatic techniques \cite{shavitt2009many} to arrive at the results.  In particular, Figure \ref{fig:diagrams-imsrg} shows the diagrammatic form of Eq.~\eqref{eq:flow0}, Eq.~\eqref{eq:flow1}, and Eq.~\eqref{eq:flow2}.

The commutator in the flow equations Eq.~\eqref{eq:imsrgode} ensures that the evolved state $\hat U(s) \ket{\Phi}$ consists of \emph{linked diagrams} only \cite{shavitt2009many}.  This indicates that IM-SRG is a size-extensive \cite{ISI:A1981MN73700014} method by construction, even if the operators are truncated.

An accurate and robust ODE solver is required to solve Eq.~\eqref{eq:imsrgode}.  In particular, the solver must be capable of handling the stiffness that often arises in such problems.  For our numerical experiments, we used a high-order ODE solver algorithm by L.\ F.\ Shampine and M.\ K.\ Gordon \cite{shampine1975computer}, which is a multistep method based on the implicit Adams predictor-corrector formulas.  Its source code is freely available \cite{odesolver}.

@@

The effective Hamiltonian in Eq.~\eqref{eq:cc_heff0} can be rewritten with commutators according to the Baker--Campbell--Hausdorff expansion as,
\begin{align*}
  \bar{H}_{\mathrm{CC}} = \hat{H} + [\hat{H}, \hat{T}] + \frac{1}{2} [[\hat{H}, \hat{T} ], \hat{T}] + \frac{1}{3!} [[[\hat{H}, \hat{T}], \hat{T}], \hat{T}] + \frac{1}{4!} [[[[\hat{H}, \hat{T}], \hat{T}], \hat{T}], \hat{T}],
\end{align*}
which terminates at four-nested commutators due to the two-body nature of the interaction.  Like with IM-SRG, this commutator expression ensures that coupled cluster is size-extensive and contains only connected terms.  In addition, because $\hat{T}$ is an excitation operator, terms of the form $\hat{T} \hat{H}$ are disconnected and thus vanish \cite{shavitt2009many}.  Therefore the coupled cluster effective Hamiltonian can be further reduced to
\begin{align} \label{eq:cc_heff1}
  \bar{H}_{\mathrm{CC}} = \bigl(\hat{H}e^{\hat{T}}\bigr)_{\mathrm{c}},
\end{align}
where the subscript ``$\mathrm{c}$'' indicates that only connected terms are used.

In practice, the cluster operator $\hat{T}$ must be truncated to be computationally feasible.  In this paper, we truncate it to only single and double excitations,
\begin{align*}
  \hat{T} = \hat{T}_{1} + \hat{T}_{2}.
\end{align*}
This is known as coupled cluster singles and doubles (CCSD), with an asymptotic computational cost that scales like IM-SRG(2).  This truncation has been successfully applied to many problems in quantum chemistry \cite{RevModPhys.79.291} and nuclear physics \cite{0034-4885-77-9-096302}.  In addition, we also truncate the three-body effective Hamiltonian terms that are induced by the similarity transformation.  Figure \ref{fig:diagrams-ccsd} shows the diagrammatic representation of Eq.~\eqref{eq:cc_heff1} in CCSD.

>>>>

In this work we will mainly work with the formalism of second quantization, see for example Ref.\ \cite{shavitt2009many}.  To properly describe fermions, we introduce the annihilation and creation operators $\hat a_p$ and $\hat a_p^\dagger$. These operators must satisfy the anticommutation relations that encode the Pauli principle, namely
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}.
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.\ \eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.  In this paper, we use the letters $p, q, r, s, t, \ldots$ to label arbitrary single-particle \textit{states} in the basis.  For the harmonic oscillator basis, labels such as $p$ can be considered abbreviations for triplets of the form $(n, m_\ell, m_{\mathrm{s}})$.  A summation such as $\sum_p$ is then understood as a summation over every single-particle basis state $p$.  This is not to be confused with summations such as $\sum_j$ from \eqref{eq:onetwobodyhamiltonian}, which contains a label $j$ denoting an arbitrary \textit{particle}.  We thus see another benefit with second quantization: the Hamiltonian representation makes no attempt to distinguish particles at all, as expected for particles with quantum statistics.

@@

Aside from trivial cases, the HF equation is generally solved using an iterative algorithm.  We begin with an initial guess $\bm{C}^{(i)}$, which is fed into Eq.\ \eqref{eq:fock} to produce the Fock matrix.  This is then used in Eq.\ \eqref{eq:hartreefock}, which leads to a standard eigenvalue problem from which $\bm{C}^{(i + 1)}$ arise as the matrix of eigenvectors and $\bm{\varepsilon}^{(i + 1)}$ as the vector of eigenvalues.  This process can be repeated indefinitely until $\bm{C}$ approaches a fixed point.  While in theory it is possible for the solution to never reach a fixed point, or that it may require a unfeasibly large number of iterations, in practice this naive approach can adequately provide solutions for many cases.  In other cases where it is insufficient, methods such as DIIS, Broyden's method, or even \textit{ad hoc} linear mixing can improve and accelerate convergence greatly.  Therefore, the possibility of slow or non-convergence is generally not a concern in practice.

@@

To make use of the HF solution as the reference state for future calculations, we transform the operators via Eqs.\ \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  Therefore, the post-HF methods described in the next few sections generally use the transformed operators $\hat{H}^{[1] \prime}$ and $\hat{H}^{[2] \prime}$ as inputs.  However, we will not annotate them with the prime symbols as the methods are generic and can be applied to any reference state, whether optimized by HF or not.

@@

The central theme of similarity renormalization group (SRG) methods is the application of a continuous sequence of unitary transformations on the Hamiltonian to evolve it into a band- or block-diagonal form, see for examples Refs.\ \cite{kehreinflow2006,heikophysrep2016,lnp936} for derivations and calculational details.

@@

The sequence of transformations is parameterized by a continuous variable $s$ known as the \textit{flow parameter}.  Without loss of generality, we can define $s = 0$ to be the beginning of this sequence, thus $\hat{H}(0)$ is simply the original Hamiltonian.  At any value of $s$, the evolving Hamiltonian $\hat{H}(s)$ is related to the original Hamiltonian by
\begin{align*}
  \hat{H}(s) \equiv \hat{U}(s) \hat{H}(0) \hat{U}^\dagger(s),
\end{align*}
where $U(s)$ is a unitary operator that describes the product of all such transformations since $s = 0$.  Taking the derivative with respect to $s$, we obtain:
\begin{align*}
  \frac{\D}{\D s} \hat{H}(s) = \frac{\D \hat{U}(s)}{\Ds} \hat{H}(0) \hat{U}^\dagger(s) + \hat{U}(s) \hat{H}(0) \frac{\D \hat{U}^\dagger (s)}{\D s}.
\end{align*}
If we define an operator $\hat{\eta}(s)$ as
\begin{align} \label{eq:etadefinition}
  \hat{\eta}(s) \equiv \frac{\D \hat{U}(s)}{\D s} \hat{U}^\dagger(s),
\end{align}
we find that it is antihermitian as a result of the unitarity of $\hat{U}(s)$:
\begin{align*}
  \hat{\eta}(s) + \hat{\eta}^\dagger(s)
  = \frac{\D}{\D s} \left(\hat{U}(s) \hat{U}^\dagger(s)\right)
  = 0,
\end{align*}
From this property we can derive a differential equation known as the \textit{SRG flow equation}:
\begin{align} \label{eq:imsrgode}
  \frac{\D \hat{H}(s)}{\D s} = [\hat{\eta}(s), \hat{H}(s)].
\end{align}
This equation allows $\hat{H}(s)$ to be evaluated without explicitly constructing the full transformation $\hat U(s)$.  The focus is instead shifted to the operator $\hat \eta(s)$, the \textit{generator} of the transformation.  When $\hat \eta(s)$ is \emph{multiplicatively} integrated (\textit{product integral}), the full unitary transformation $\hat U(s)$ is recovered:
\begin{align} \label{eq:etaintegral}
  \hat U(s')
  = \lim_{\Delta s \to 0} \prod_{i = 1}^n \mathrm{e}^{\hat{\eta}(s_i) \Delta s},
\end{align}
where $s_i \equiv i \Delta s$, $n \equiv \lfloor s' / \Delta s \rfloor$, and $\lfloor x \rfloor$ denotes the \textit{floor} of $x$.  This is the formal solution to the linear differential equation Eq.\ (\ref{eq:etadefinition}).  The product integral in Eq.\ (\ref{eq:etaintegral}) may also be reinterpreted as \textit{$s$-ordering} \cite{reimann2013quantum} in analogy to time-ordering from quantum field theory.

@@

The SRG flow equation Eq.\ \eqref{eq:imsrgode} can be solved in the second quantization formalism described in Section \ref{subsec:modelHamiltonian}, where field operators are defined with respect to the physical vacuum state.  However, since the basis of a many-body problem grows factorially with the number of particles and the size of the model space, the applicability of the naive (free-space) SRG method is restricted to comparatively small systems.  A more practical approach is to perform the evolution \textit{in medium}, i.e.\ using a many-body Slater determinant as a reference \cite{kehrein2006flow}.  This gives rise to the IM-SRG method \cite{bogner2012,heikophysrep2016,lnp936}.

The second-quantized operators in Eq.\ \eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1, i_2, \ldots, i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators.  The normal ordering of the quasiparticle field operators is straightforward to compute.  What is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $|\Phi\rangle$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &=
  \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p, q}^{}.
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_{i} n_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} n_i n_j \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_{i} n_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix $F_{p q}$.  The operator $\hat H$ in Eq.\ \eqref{eq:normordhamiltonian} is completely equivalent to $\hat H$ in Eq.\ \eqref{eq:vacuumhamiltonian}.  The only difference is that the meaning of a ``$k$-body operator'' has been redefined with respect to the reference state $\Phi$ rather than vacuum state $\ket{}$, causing matrix elements to be reshuffled among the $k$-body operators.  Where this makes a critical difference is when the operator expressions are \emph{truncated}, i.e.\ some $k$-body operators may be discarded from the computation for efficiency reasons.  Normal-ordering can drastically alter the importance of the $k$-body operators with respect to each other.

@@

By expanding the commutator in Eq.\ \eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H) \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \mathcal{A}_{A, B} D^{[k]}(A, B) \text{ for any $k$, $A$, $B$}, \label{eq:commutmxe} \\
  \mathcal{A}_{x, y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr) \text{ for any $x$, $y$, $f$},
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} n_i \bar{n}_a A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} n_i n_j \bar{n}_a \bar{n}_b A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p, q}(A, B)
  &\equiv
    \sum_{r} A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} n_i n_j \bar{n}_a A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} n_i \bar{n}_a \bar{n}_b A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} n_i \bar{n}_a \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right)
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    \tilde{\mathcal{A}}_{p, q} \tilde{\mathcal{A}}_{r s}
    \sum_{i a} n_i \bar{n}_a A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    \notag \\
  &\hphantom{=}
    + \frac{1}{2} \sum_{i j} n_i n_j A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} \bar{n}_a \bar{n}_b A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + \sum_{t} \left(
    \tilde{\mathcal{A}}_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \tilde{\mathcal{A}}_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}.
    \right)
    \label{eq:flow2}
\end{align}

\begin{figure}
\includegraphics{fig-diagrams-imsrg.pdf}
\caption{Diagrammatic representation of $D^{[k]}(\circ, \bullet)$ of Eq.\ \eqref{eq:flow0}, Eq.\ \eqref{eq:flow1}, and Eq.\ \eqref{eq:flow2}, where open circles represent $A$ and filled circles represent $B$.  The two-body vertices are implicitly antisymmetric (i.e.\ \textit{Hugenholtz diagrams} \cite{HUGENHOLTZ1957481}, as described in \cite{shavitt2009many} \S 4.4.3).  Internal directed lines (bound/dummy indices) that point upward denote summations over unoccupied states, whereas lines that point downward denote summations over occupied states.  External undirected lines denote unconstrained free indices.  In IM-SRG(2), the three-body diagrams are not included.}
\label{fig:diagrams-imsrg}
\end{figure}

The equations can be derived directly from the anticommutation relations of the field operators, but it is more convenient to use Wick's theorem \cite{PhysRev.80.268} or, even more efficiently, diagrammatic techniques \cite{shavitt2009many} to arrive at the results.  In particular, Figure \ref{fig:diagrams-imsrg} shows the diagrammatic form of Eq.\ \eqref{eq:flow0}, Eq.\ \eqref{eq:flow1}, and Eq.\ \eqref{eq:flow2}.

The commutator in the flow equations Eq.\ \eqref{eq:imsrgode} ensures that the evolved state $\hat U(s) \ket{\Phi}$ consists of \emph{linked diagrams} only \cite{shavitt2009many}.  This indicates that IM-SRG is a size-extensive \cite{ISI:A1981MN73700014} method by construction, even if the operators are truncated.

An accurate and robust ODE solver is required to solve Eq.\ \eqref{eq:imsrgode}.  In particular, the solver must be capable of handling the stiffness that often arises in such problems.  For our numerical experiments, we used a high-order ODE solver algorithm by L.\ F.\ Shampine and M.\ K.\ Gordon \cite{shampine1975computer}, which is a multistep method based on the implicit Adams predictor-corrector formulas.  Its source code is freely available \cite{odesolver}.

@@

The effective Hamiltonian in Eq.\ \eqref{eq:cc_heff0} can be rewritten with commutators according to the Baker--Campbell--Hausdorff expansion as,
\begin{align*}
  \bar{H}_{\mathrm{CC}} = \hat{H} + [\hat{H}, \hat{T}] + \frac{1}{2} [[\hat{H}, \hat{T} ], \hat{T}] + \frac{1}{3!} [[[\hat{H}, \hat{T}], \hat{T}], \hat{T}] + \frac{1}{4!} [[[[\hat{H}, \hat{T}], \hat{T}], \hat{T}], \hat{T}],
\end{align*}
which terminates at four-nested commutators due to the two-body nature of the interaction.  Like with IM-SRG, this commutator expression ensures that coupled cluster is size-extensive and contains only connected terms.  In addition, because $\hat{T}$ is an excitation operator, terms of the form $\hat{T} \hat{H}$ are disconnected and thus vanish \cite{shavitt2009many}.  Therefore the coupled cluster effective Hamiltonian can be further reduced to
\begin{align} \label{eq:cc_heff1}
  \bar{H}_{\mathrm{CC}} = \bigl(\hat{H}e^{\hat{T}}\bigr)_{\mathrm{c}},
\end{align}
where the subscript ``$\mathrm{c}$'' indicates that only connected terms are used.

In practice, the cluster operator $\hat{T}$ must be truncated to be computationally feasible.  In this paper, we truncate it to only single and double excitations,
\begin{align*}
  \hat{T} = \hat{T}_{1} + \hat{T}_{2}.
\end{align*}
This is known as coupled cluster singles and doubles (CCSD), with an asymptotic computational cost that scales like IM-SRG(2).  This truncation has been successfully applied to many problems in quantum chemistry \cite{RevModPhys.79.291} and nuclear physics \cite{0034-4885-77-9-096302}.  In addition, we also truncate the three-body effective Hamiltonian terms that are induced by the similarity transformation.  Figure \ref{fig:diagrams-ccsd} shows the diagrammatic representation of Eq.\ \eqref{eq:cc_heff1} in CCSD.

\begin{figure}
  \includegraphics{fig-diagrams-ccsd.pdf}
  \caption{Diagrammatic representation of $\bar{H}$ of Eq.\ \eqref{eq:cc_heff1}, where open circles represent the excitation cluster operators $\hat{T}_{1}$ and $\hat{T}_{2}$, and filled circles represent the two-body interaction $\hat{H}^{[2]}$.  As before, the diagrams are implicitly antisymmetrized (Hugenholtz diagrams).  Lines connected to $\hat{T}$ are always directed upward because they represent an excitation operator while the directions of external lines connected to $\hat{H}^{[2]}$ are unconstrained.}
  \label{fig:diagrams-ccsd}
\end{figure}
